Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-08 numerical.tex
updating

Fig. \ref{bcrit_compare} shows the growth rate $\nu$ of the
fundamental VSI mode as a function of $\beta$ for $\khat=30$. This is
approximately the perturbation wavenumber observed in the fiducial
numerical simulation of \cite{nelson13}. We also plot the growth rate 
obtained by numerically solving the dispersion relation 
Eq. \ref{relax_disp} and the estimate for the upper limit of
the thermal relaxation timescale $\beta_\mathrm{crit}$ given by
Eq. \ref{iso_vsi_cond}. For the present disk parameters
$\beta_\mathrm{crit}=0.125$. We find good agreement between the
numerical and analytical growth rates, as well as the critical
relaxation timescale.

 \begin{figure}
   \includegraphics[width=\linewidth]{figures/bcrit_compare} 
   \caption{Growth rate of the fundamental VSI mode with $\khat=30$ as
     a function of the thermal relaxation timescale $\beta$. The disk
     parameters are $\Gamma=1.011$, $\gamma=1.4$, and 
     $(p,q,\epsilon)=(-1.5,-1,0.05)$. The numerical (analytical) growth
     rates are show as solid (dotted) lines, while the vertical line
     is the maximum thermal timescale $\beta_\mathrm{crit}$ for the
     fundamental mode according to Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare}}   
 \end{figure}

@@

We visualize the effect of thermal relaxation in 
Fig. \ref{relax_eigenW_num} by comparing the eigenfunction $W$ for
$\beta=0.01$ and $\beta=0.125$. The latter value is close to the critical
value beyond which the fundamental isothermal VSI is stabilized
(\S\ref{iso_vsi_beta_crit}). Hence the latter growth rate is smaller
than the former by a factor $\sim 20$. Increasing $\beta$ restricts 
the region in which $W\sim z$ closer to the midplane, with oscillatory
behavior emerging at the boundaries, where buoyancy first becomes
important. % We also find this trend for fixed $\beta$ but increasing
% $\khat$.

@@

\subsubsection{$\gamma=\Gamma \neq 1$, $N_z^2\equiv 0$}
Consider first the case $\gamma=1.4$ so that there is no vertical
entropy gradient. We plot in Fig. the growth rates for the
fundamental VSI for $\beta\in[10^{-2},10^{2}]$. We find growth rates
do not reach zero, implying instability regardless of the thermal
relaxation timescale if $\gamma=\Gamma$. This result is consistent
with \cite{nelson13}.  Note that for $\beta\to\infty$ (adiabatic
perturbations) and $\gamma=\Gamma$, the disk can be unstable according
to the Solberg-Hoiland criterion  Eq. \ref{solberg2}.

Growth rates for $\beta = 10^{-2}$ and
$\beta=10^{2}$ behave similarly. This is not surprising because
the linearized energy equation (Eq. \ref{lin_energy}) implies that 
\begin{align}
  W = \frac{\left(\Gamma^{-1} -
      \ii\hat{\sigma}\beta\right)}{\left(1-\ii\hat{\sigma}\beta\right)}
  Q \text{ for } \gamma\equiv \Gamma.   
\end{align}
Thus, $W \to Q/\Gamma$ as $\beta\to0$ and $W\to Q$ as
$\beta\to\infty$. Since $\Gamma=O(1)$, these two limits are expected
to produce qualitatively similar results. Interestingly,
Fig. \ref{growth_vnoniso1} show that growth rates 
maximize for $\beta=O(1)$. However, all growth rates are similar to
order-of-magnitude for $0.1\lesssim\khat\lesssim 100$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/growth_vnoniso1} 
  \caption{Numerically-determined growth rates of the fundamental VSI
    mode in neutrally-buoyant ($N_z^2\equiv0$), vertically non-isothermal disks
    ($\Gamma=\gamma=1.4$), with $(p,q,\epsilon)=(0,-1,0.05)$, for
    a range of thermal relaxation timescales 
    $\beta$. \label{growth_vnoniso1}}     
\end{figure}

\subsubsection{$\gamma\neq\Gamma \neq 1$, $N_z^2\geq0$}
%gamma = 2.0, Gamma = 1.4, vary bcool 
Finally, we introduce buoyancy in a vertically non-isothermal disk by
setting $\gamma=2$. This scenario was not considered by  
\cite{nelson13}, but is similar in spirit with vertically isothermal 
disks evolved with $\gamma>1$, as considered in the previous section. 
Fig. \ref{growth_vnoniso2} plots the fundamental VSI growth rates for
this case. We observe the same trend as for vertically isothermal
disks with $N_z^2\geq 0$ (Fig. \ref{relax_growth_num}), namely growth
rates reach zero at finite $\khat$. From Fig.\ref{growth_vnoniso2} we
expect the VSI to be strongly stabilized even for small, finite
thermal relaxation times.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/growth_vnoniso2}
  \caption{Same as Fig. \ref{growth_vnoniso1}, but with $\gamma=2$ so that
    $N_z^2\geq0$. \label{growth_vnoniso2}}
\end{figure}

>>>>

Fig. \ref{bcrit_compare1} shows the growth rate $\nu$ of the
fundamental VSI mode as a function of $\beta$ for $\khat=30$. This is
approximately the perturbation wavenumber observed in the fiducial
numerical simulation of \cite{nelson13}. We also plot the growth rates
obtained by numerically solving the dispersion relation 
Eq. \ref{relax_disp} and the upper limit to 
the thermal relaxation timescale $\beta_\mathrm{crit}$ given by
Eq. \ref{iso_vsi_cond}. For the present disk parameters
$\beta_\mathrm{crit}=0.125$. We find good agreement between the
numerical and analytical growth rates, as well as the critical
relaxation timescale. Growth rates are insensitive to $\beta$ for
$\beta\lesssim 0.05$, beyond which growth rates rapidly decrease.

%implications for nonlinear sims?

 \begin{figure}
   \includegraphics[width=\linewidth]{figures/bcrit_compare} 
   \caption{Growth rate of the fundamental VSI mode with $\khat=30$ as
     a function of the thermal relaxation timescale $\beta$. The disk
     parameters are $\Gamma=1.011$, $\gamma=1.4$, and 
     $(p,q,\epsilon)=(-1.5,-1,0.05)$. The numerical (analytical) growth
     rates are show as solid (dotted) lines, while the vertical line
     is the maximum thermal timescale $\beta_\mathrm{crit}$ for the
     fundamental mode according to Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

@@

We visualize the effect of thermal relaxation in 
Fig. \ref{relax_eigenW_num} by comparing the eigenfunction $W$ for
$\beta=0.01$ and $\beta=0.1$. The latter value is close to the
critical value beyond which the fundamental VSI is stabilized
(\S\ref{iso_vsi_beta_crit}). Increasing $\beta$
restricts  the region in which $W\sim z$ closer to the midplane, with
oscillatory behavior emerging at the boundaries, where buoyancy first
becomes important.

% We also find this trend for fixed $\beta$ but
% increasing 
% $\khat$.

@@

% \subsubsection{$\gamma=\Gamma \neq 1$, $N_z^2\equiv 0$}
% Consider first the case $\gamma=1.4$ so that there is no vertical
% entropy gradient. We plot in Fig. the growth rates for the
% fundamental VSI for $\beta\in[10^{-2},10^{2}]$. We find growth rates
% do not reach zero, implying instability regardless of the thermal
% relaxation timescale if $\gamma=\Gamma$. This result is consistent
% with \cite{nelson13}.  Note that for $\beta\to\infty$ (adiabatic
% perturbations) and $\gamma=\Gamma$, the disk can be unstable according
% to the Solberg-Hoiland criterion  Eq. \ref{solberg2}.

% Growth rates for $\beta = 10^{-2}$ and
% $\beta=10^{2}$ behave similarly. This is not surprising because
% the linearized energy equation (Eq. \ref{lin_energy}) implies that 
% \begin{align}
%   W = \frac{\left(\Gamma^{-1} -
%       \ii\hat{\sigma}\beta\right)}{\left(1-\ii\hat{\sigma}\beta\right)}
%   Q \text{ for } \gamma\equiv \Gamma.   
% \end{align}
% Thus, $W \to Q/\Gamma$ as $\beta\to0$ and $W\to Q$ as
% $\beta\to\infty$. Since $\Gamma=O(1)$, these two limits are expected
% to produce qualitatively similar results. Interestingly,
% Fig. \ref{growth_vnoniso1} show that growth rates 
% maximize for $\beta=O(1)$. However, all growth rates are similar to
% order-of-magnitude for $0.1\lesssim\khat\lesssim 100$.

% \begin{figure}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
%   0cm]{figures/growth_vnoniso1} 
%   \caption{Numerically-determined growth rates of the fundamental VSI
%     mode in neutrally-buoyant ($N_z^2\equiv0$), vertically non-isothermal disks
%     ($\Gamma=\gamma=1.4$), with $(p,q,\epsilon)=(0,-1,0.05)$, for
%     a range of thermal relaxation timescales 
%     $\beta$. \label{growth_vnoniso1}}     
% \end{figure}

% \subsubsection{$\gamma\neq\Gamma \neq 1$, $N_z^2\geq0$}
% %gamma = 2.0, Gamma = 1.4, vary bcool 
% Finally, we introduce buoyancy in a vertically non-isothermal disk by
% setting $\gamma=2$. This scenario was not considered by  
% \cite{nelson13}, but is similar in spirit with vertically isothermal 
% disks evolved with $\gamma>1$, as considered in the previous section. 
% Fig. \ref{growth_vnoniso2} plots the fundamental VSI growth rates for
% this case. We observe the same trend as for vertically isothermal
% disks with $N_z^2\geq 0$ (Fig. \ref{relax_growth_num}), namely growth
% rates reach zero at finite $\khat$. From Fig.\ref{growth_vnoniso2} we
% expect the VSI to be strongly stabilized even for small, finite
% thermal relaxation times.

% \begin{figure}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
%   0cm]{figures/growth_vnoniso2}
%   \caption{Same as Fig. \ref{growth_vnoniso1}, but with $\gamma=2$ so that
%     $N_z^2\geq0$. \label{growth_vnoniso2}}
% \end{figure}
